---
# bibliography: here::here(OffshoreHabitat.bib)
# csl: plos.csl
# fontsize: 10pt
# geometry: left=2cm, right=2cm, top=2cm, bottom=3cm, footskip = .5cm
# link-citations: yes
# output:
#   bookdown::pdf_document2:
#     includes:
#        in_header: latex/header.tex
#     keep_tex: true
#   bookdown::html_document2:
#     toc: true
#     toc_float: true
#     code_fold: hide
#   bookdown::word_document2:
#     toc: true
# subparagraph: yes
# urlcolor: blue
---

```{r setup, include=FALSE}

# knitr::opts_chunk$set(echo = FALSE,
#                       message = FALSE,
#                       #dev = "cairo_pdf",
#                       warning = FALSE,
#                       fig.width = 4,
#                       fig.asp = 0.45,
#                       fig.align = 'center'
#                       ) #allows for inserting R code into captions

```

### Offshore Habitat (new)

**Description:**

This element is applied at the species level.

Offshore habitat, defined here as all habitat outside of the estuary
and beyond the immediate coastal/nearshore areas, supports all life
stages of many Council managed species, and is changing in quality and
quantity due to multiple stressors from climate to other ocean uses such
as offshore wind development. This element evaluates risk of achieving
OY due to changes in offshore habitat quality and quantity.

**Definition:**

Risk of not achieving OY due to changing offshore habitat. The rationale
is that multiple drivers of offshore habitat change, including ocean
industrialization, are included in this definition.

**Bluefish Indicators:**

Numerous indicators were developed for the Bluefish ESP (Tyrell et al., 2022). Some of those indicators are also relevant to offshore habitat. The indicators pertain to Distribution, Climate, and Natural Mortality. 

Distribution indicators are generated by a VAST model from NMFS bottom trawl data to assess shifts in the center of gravity of three different size classes of Bluefish in Northings and Eastings. Changes in Bluefish distribution could inform state quotas, impact location and timing of spawning, and impact the location of capture.

```{r bfcog, fig.cap="Bluefish Center of Gravity by size class in Northings (km) and Eastings (km)", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_cog.R'))
plot_bf_cog()
```

Climate indicators are generated from OISST satellite data from the Physical Sciences Laboratory. These indicators evaluate: the first and last day of the year when the mean temperature of the Central Atlantic bluefish region is above 18 C, the number of days when >75% of the region is above 18 C, the mean proportion of the region below 18 C, between 18-25.6 C, and above 25.6 C in July, and the mean cross-shore and along shore wind speeds in the region in April and May. Upper and lower thermal thresholds for spawning have been identified as 25.6 and 18 C respectively (Norcross et al., 1974). The date when these temperature thresholds are reached and the proportion of habitat within this thermal tolerance may impact the spatial temporal extent of the spawning season and therefore the number of individuals spawned.Along shore and cross-shore winds may help transport larvae and juveniles inshore impacting recruitment(Norcross et al., 1974).

```{r bf18, fig.cap="First, last, and number of days when the mean temperature of the central Atlantic region is warmer than 18 C. Dashed lines depict the time series mean. Trend lines depict statistically significant trends.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_climate.R'))
plot_bf_climate()
```

```{r bfjp, fig.cap="Proportion of the central Atlantic below, within, and above the thermal tolerance of Bluefish. Dashed lines depict time series means. Trend lines depict statistically significant trends", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_july_proportion.R'))
plot_bf_july_proportion()
```
```{r bfw, fig.cap="Mean cross and along shore wind in the central Atlantic in April and May. Dashed lines depict time series means", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_wind_indicators.R'))
plot_bf_wind_indicators()
```


Natural mortality indicators assessed Mako shark B/Bmsy from the International Commission for the Conservation of Atlantic Tunas (ICCAT) stock assessment and the relative condition of three size classes of Bluefish aggregated from multiple state and federal surveys as described by Truesdell et al. (2022). Large predators, including Mako sharks, are known to prey on Bluefish and impact natural mortality (Shepherd and Packer 2006, Wood and Wetherbee 2009). Bluefish with poor condition may be more susceptible to predation and disease, resulting in higher natural mortality.

```{r mako, fig.cap="Mako shark B/Bmsy. Dashed line depicts time series mean. Trend line depicts statistically significant trend", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_mako.R'))
plot_bf_mako()
```

```{r bfcond, fig.cap="Spring and Fall relative condition of three different size classes of Bluefish. Dashed lines depict time series means. Trend lines depict statistically significant trends.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_condition.R'))
plot_bf_condition()
```


**Black sea bass Indicators:**

Black sea bass indicators are generated from the Black Sea Bass ESP (Tabandera et al., 2024). The indicators also pertain to Offshore Habitat.

A mean winter (February and March) bottom temperature indicator was developed from observed ocean temperature data from the Hydrographic Database maintained by the NEFSC Oceans and Climate Branch. These are composed of CTD measurements concurrent with bottom trawl surveys and other deployed instruments. Quality assurance and control are performed on these sources to produce area weighted regional mean values as described in Fratantoni et al. (2013). Mean values were computed for each survey and a reference annual cycle (fit to observations from 1981-2010) was removed to produce anomalies. The values in the dataset are anomalies plus expected mean value for each location and time of year based on this reference period. To assess winter conditions, February and March data were averaged on the regional scale. January was excluded due to low in-situ data coverage. Mean winter bottom temperature was explored as an indicator of overwinter survival. Black sea bass have an incipient lethal temperature threshold at 6°C, below which feeding ceases and temperature-induced death is the primary cause of mortality (Younes et al. 2020).

```{r bsbwbt, fig.cap="Mean winter (February and March) bottom temperature. Dashed line depicts lower thermal tolerance. Trend line depicts statistically significant trend.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_winter_bottom_temp.R'))
plot_winter_bottom_temp()
```


A high winter (February and March) shelf water volume indicator was developed from data from the Hydrographic Database mainteind by the NEFSC Oceans and Climate Branch. Shelf water is defined as all water having salinity <34 psu (the isohaline typically associated with the shelf-slope front in this region). The methods to compute this time series are described in Fratantoni et al. (2012). High shelf water volume was explored as an indicator of overwinter mixing between the northern and southern subregions. Miller et al. (2016) identified a negative impact on BSB catch when shelf water volume exceeded 4000 km3, which indicates that higher shelf water volumes could result in lower suitable habitat for BSB, and proposed that years with shelf water volume >4000 km3 could be used to identify possible mixing between subregions.

```{r bsbwsvol, fig.cap="Annual mean winter (February and March) shelf water volume. Dashed line depicts the proposed threshold above which overwinter mixing is more likely.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_winter_shw_vol.R'))
plot_winter_shw_vol()
```


**Golden tilefish Indicators:**

Numerous indicators were developed for the Golden Tilefish ESP (Salois et al., 2024). Some of those indicators are also relevant to offshore habitat. The indicators pertain to habitat condition and physical oceanography.

Habitat condition indicators assessed the cold pool index, cold pool spatial extent index, cold pool persistence index, and the mean bottom temperature and salinity of the spatial extent in which Golden tilefish have been captured by the NMFS bottom trawl survey. The cold pool index indicators are generated from bias-corrected ROMS-NWA and GLORYS data. The lower bound of temperature preference for eggs and larvae and juveniles is between 8-13 °C. If the cold pool has a larger extent or persists for a longer time, environmental conditions are not optimal for early life history stages and recruitment success. Bottom temperature and salinity indicators were generated using data from GLORYS12 (E.U. Copernicus Marine Service Information, 2023), a global eddy-resolving physical ocean and sea ice model (1993-2023) with high resolution (1/12°). Optimal bottom temperature for golden tilefish is between 9-14 °C (Grimes and Turner, 1999). Temperatures outside this range may impact spawning location/success and recruitment success. Optimal bottom salinity for adult golden tilefish is 33-36 psu for juveniles and adults (Steimle et al. 1999). Salinities outside this range may affect spawning and recruitment success.

```{r gtfcp, fig.cap="Cold pool index, cold pool persistence index, and cold pool spatial extent index based on bias-corrected ROMS-NWA (open circles) and GLORYS (closed circles)", echo=FALSE}
ecodata::plot_cold_pool()
```


```{r gtfbt, fig.cap="Monthly weighted mean bottom temperature (at 78 m depth) of NMFS bottom trawl stat areas in which Golden tilefish has been captured.Dashed lines depict the upper and lower thermal tolerance of the species. Confidence interval shading represents +/- 2 standard deviations around the weighted mean", echo=FALSE}
source(here::here('offshore_habitat/R/plot_gtf_bottom_temp.R'))
plot_gtf_bottom_temp()
```

```{r gtfbs, fig.cap="Annual average bottom salinity (at 78 m depth) of NMFS bottom trawl stat areas in which Golden tilefish has been captured.Dashed lines depict the upper and lower thermal tolerance of the species. Confidence interval shading represents +/- 2 standard deviations around the average", echo=FALSE}
source(here::here('offshore_habitat/R/plot_gtf_bottom_sal.R'))
plot_gtf_sal()
```


Physical oceanography indicators assessed the sea surface temperature (SST), gulf stream index (GSI), and shelf water volume, temperature, and salinity. SST was derived from the Advanced Clear Sky Processor for Ocean (ACSPO) system. Sea surface temperatures can be used as a proxy for water mass movement (Fisher et al., 2014), which may cause displacement or mortality of eggs and larvae. The GSI is the degrees latitude above the average Gulf Stream position based on ocean temperature at 200 m (15 C) depth between 55°W to 75°W. Further details are available at https://noaa-edab.github.io/ecodata/reference/gsi.html . A more northerly position of the Gulf Stream (positive Gulf Stream Index) pushes warm water onto the shelf, creating more suitable habitat for Golden tilefish. Shelf water volume, temperature, and salinity were derived from the Hydrographic Database maintained by the NEFSC Oceans and Climate Branch, full details described in Fratantoni et al. (2012). Shelf water physical oceanography indicators were compare to tolerance thresholds from the literature to assess changes in habitat suitability.

```{r ltsst, fig.cap="Northeast US annual sea surface temperature (SST, black), with increasing trend (orange).", echo=FALSE}
ecodata::plot_long_term_sst()
```
```{r gsi, fig.cap="Gulf stream index. Dashed line depicts time series mean. Trend line depicts statistically significant trend.", echo=FALSE}

ecodata::plot_gsi()

```

```{r gtfshwvol, fig.cap="Annual mean shelf water volume. Dashed line depicts time series mean", echo=FALSE}
source(here::here('offshore_habitat/R/plot_shw_vol.R'))
plot_shw_vol()
```

```{r gtfshwt, fig.cap="Annual mean shelf water temperature. Dashed line depicts time series mean. Trend line depicts statistically significant trend.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_shw_temp.R'))
plot_shw_temp()

```

```{r gtfshws, fig.cap="Annual mean shelf water salinity. Dashed line depicts time series mean.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_shw_sal.R'))
plot_shw_sal()

```





**Risk criteria:**

  Bluefish
  -----------------------------------------------------------------------
  *Risk Level*    *Definition*
  --------------- -------------------------------------------------------
  Low             No indicators suggest negative impact to OY

  Low-Moderate    Either climate indicators or natural mortality 
                  indicators suggest negative impact to OY but 
                  distribution indicators suggest no impact.

  Moderate-High   Distribution indicators suggest negative impacts 
                  to OY or Both climate and natural mortality 
                  indicators suggest negative impacts to OY.

  High            Distribution, climate, and natural mortality 
                  indicators all suggest negative impacts to OY.
  -----------------------------------------------------------------------
  
  Black sea bass  
  -----------------------------------------------------------------------
  *Risk Level*    *Definition*
  --------------- -------------------------------------------------------
  Low             No indicators suggest negative impact to OY

  Low-Moderate    Indicators suggest increase in overwinter mixing
                  but no change or increase in overwinter survival

  Moderate-High   Indicators suggest decrease in overwinter survival
                  but no change or decrease in overwinter mixing

  High            Overwinter survival decreasing, overwinter mixing 
                  increasing
  -----------------------------------------------------------------------
  
  Golden tilefish  
  -----------------------------------------------------------------------
  *Risk Level*    *Definition*
  --------------- -------------------------------------------------------
  Low             No indicators suggest negative impact to OY

  Low-Moderate    Physical oceanography indicators suggest negative
                  impacts to OY but Habitat condition indicators 
                  suggest no impact or positive impact

  Moderate-High   Habitat condition indicators suggest negative 
                  impacts to OY but Physical oceanography indicators
                  suggest no impact or positive impact

  High            Both Physical oceanography and Habitat condition 
                  indicators suggest negative impact to OY
  -----------------------------------------------------------------------

**Risk Assessment**

Bluefish: A mix in indicator trends results in a moderate-high ranking for this element. Distribution indicators suggest a negative impact to OY as there are significant trends in Medium (30.3-50 cm) and Large (>50 cm) center of gravity both in Eastings and Northings. Climate indicators show mixed signals. Significant trends in some indicators suggest lengthening spawning season and potentially greater spawning success (increase in last day warmer than 18 C, increase in number of days warmer than 18 C, decrease in proportion of the central Atlantic colder than 18 C in July). One climate indicator suggests a growing proportion of the region is above the upper thermal tolerance of the species which could lead to decreased spawning success (increasing trend in proportion of the central Atlantic warmer than 25.6 C in July).Natural mortality indicators suggest a positive impact on OY. Mako shark B/Bmsy has a negative trend, indicative of decreased predation and natural mortality of Bluefish. There is a positive trend in Large Bluefish relative condition in the spring which could have positive impacts on survival and reproductive success.

Black sea bass: No indicators suggest a negative impact on OY, resulting in a low ranking for this element. Mean winter bottom temperature is increasing away from the lower temperature threshold of the species which has a positive impact on recruitment. There is no trend in winter mean shelf volume and the majority of the time series is below the volume threshold for increased risk of overwinter mixing between sub regions.

Golden tilefish:No indicators suggest a negative impact on OY, resulting in a low ranking for this element. Trends in the cold pool index and spatial extent suggest warmer conditions and a smaller cold pool which will likely have positive impacts on Golden tilefish as the average cold pool temperature is below the lower thermal tolerance of the species. Increasing trends in SST, GSI, and mean shelf water temperature will similarly result in increases in thermal habitat and therefore survival rates.