---
bibliography: OffshoreHabitat.bib
csl: plos.csl
fontsize: 10pt
geometry: left=2cm, right=2cm, top=2cm, bottom=3cm, footskip = .5cm
link-citations: yes
output:
  bookdown::pdf_document2:
    includes:
       in_header: latex/header.tex
    keep_tex: true
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_fold: hide
  bookdown::word_document2:
    toc: true
subparagraph: yes
urlcolor: blue
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      #dev = "cairo_pdf",
                      warning = FALSE,
                      # fig.width = 4,
                      # fig.asp = 0.45,
                      fig.align = 'center'
                      ) #allows for inserting R code into captions

```

### Offshore Habitat (new)

**Description:**

This element is applied at the species level.

Offshore habitat, defined here as all habitat outside of the estuary and beyond the immediate coastal/nearshore areas, supports all life stages of many Council managed species, and is changing in quality and quantity due to multiple stressors from climate to other ocean uses. This element evaluates risk of achieving OY due to changes in offshore habitat quality and quantity.

**Definition:**

Risk of not achieving OY due to changing offshore habitat.

**Bluefish Indicators:**

Numerous indicators were developed for the Bluefish ESP [@tyrell_bluefish_2022]. Some of those indicators are relevant to offshore habitat. The indicators pertain to Distribution, Climate, and Natural Mortality. 

Changes in Bluefish distribution could inform state quotas, impact location and timing of spawning, and impact the location of capture. Distribution indicators are generated by a VAST model from NMFS bottom trawl data to assess shifts in the center of gravity of three different size classes of Bluefish in Northings and Eastings. 

```{r bfcog, fig.cap="Bluefish Center of Gravity by size class in Northings (km) and Eastings (km)", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_cog.R'))
plot_bf_cog()
```

Climate indicators use identified thermal tolerance of Bluefish spawning activity to assess changes in the spatiotemporal availability of suitable spawning habitat. Upper and lower thermal thresholds for spawning have been identified as 25.6 and 18 C respectively [@norcross_development_1974]. Additional climate indicators were developed to assess the impact of wind on early life history. Along shore and cross-shore winds may help transport larvae and juveniles inshore impacting recruitment[@norcross_development_1974]. Temperature-related cliamte indicators are generated from OISST satellite data from the Physical Sciences Laboratory. Wind-related climate indicators were generated from monthly mean wind speed data downloaded from the Physical Sciences Laboratory. Climate indicators evaluate: the first and last day of the year when the mean temperature of the Central Atlantic Bluefish region is above 18 C, the number of days when >75% of the region is above 18 C, the mean proportion of the region below 18 C, between 18-25.6 C, and above 25.6 C in July, and the mean cross-shore and along shore wind speeds in the region in April and May.

```{r bf18, fig.cap="First, last, and number of days when the mean temperature of the central Atlantic region is warmer than 18 C. Dashed lines depict the time series mean. Trend lines depict statistically significant trends.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_climate.R'))
plot_bf_climate()
```

```{r bfjp, fig.cap="Proportion of the central Atlantic below, within, and above the thermal tolerance of Bluefish. Dashed lines depict time series means. Trend lines depict statistically significant trends", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_july_proportion.R'))
plot_bf_july_proportion()
```
```{r bfw, fig.cap="Mean cross and along shore wind in the central Atlantic in April and May. Dashed lines depict time series means", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_wind_indicators.R'))
plot_bf_wind_indicators()
```


Natural mortality indicators consider changes in offshore habitat which may impact this management parameter and therefore OY. Large predators, including Mako sharks, are known to prey on Bluefish and impact natural mortality [@shepherd_essential_2006] [@wood_recalculated_2009]. Additionally, Bluefish with poor condition may be more susceptible to predation and disease, resulting in higher natural mortality. Natural mortality indicators assessed Mako shark B/Bmsy from the International Commission for the Conservation of Atlantic Tunas (ICCAT) stock assessment and the relative condition of three size classes of Bluefish aggregated from multiple state and federal surveys [@truesdell_wp_2022].

```{r mako, fig.cap="Mako shark B/Bmsy. Dashed line depicts time series mean. Trend line depicts statistically significant trend", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_mako.R'))
plot_bf_mako()
```

```{r bfcond, fig.cap="Spring and Fall relative condition of three different size classes of Bluefish. Dashed lines depict time series means. Trend lines depict statistically significant trends.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_bf_condition.R'))
plot_bf_condition()
```


**Black sea bass Indicators:**

Indicators were developed for the Black Sea Bass ESP [@tabandera_working_2024]. The developed indicators pertain to Offshore Habitat by evaluating overwinter survival of early life stages and mixing between sub regions.

Mean winter bottom temperature was explored as an indicator of overwinter survival. Recruitment of Black sea bass (BSB) has been shown to be higher than average in warm years and below average in cold years [@tabandera_working_2024]. Additionally, BSB have a preferred temperature of >8 C [@miller_offshore_2016]. A mean winter (February and March) bottom temperature indicator was developed using observed ocean temperature data from the Hydrographic Database maintained by the NEFSC Oceans and Climate Branch. These data are composed of CTD measurements concurrent with bottom trawl surveys and other deployed instruments. Quality assurance and control were performed on these sources to produce area weighted regional mean values [@fratantoni_description_2013]. Mean values were computed for each survey and a reference annual cycle (fit to observations from 1981-2010) was removed to produce anomalies. The values in the dataset are anomalies plus expected mean value for each location and time of year based on this reference period. To assess winter conditions, February and March data were averaged on the regional scale. January was excluded due to low in-situ data coverage.

```{r bsbwbt, fig.cap="Mean winter (February and March) bottom temperature. Dashed line depicts thermal preference of >8 C. Trend line depicts statistically significant trend.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_winter_bottom_temp.R'))
plot_winter_bottom_temp()
```


High shelf water volume was explored as an indicator of overwinter mixing between the northern and southern sub regions. Overwinter mixing could lead to misallocation of removals and abundance indices between the northern and southern stock subunits leading to retrospective patterns present in previous assessments, particularly for the northern subunit. A threshold of 4000 km^3^ has been proposed, above which there is greater likelihood of mixing between sub regions [@miller_offshore_2016]. Southern MAB BSB have been shown to occupy a smaller portion of the available habitat in years with higher shelf water volume [@tabandera_working_2024]. No correlation was found between shelf water volume and the center of gravity for northern MAB [@tabandera_working_2024]. Still, it was suggested that future efforts to reallocate removals and abundance indices between the northern and southern stock subunits focus on years with high potential stock mixing. Furthermore, a negative impact on BSB catch has been identified when shelf water exceeded the proposed threshold [@miller_offshore_2016]. A high winter (February and March) shelf water volume indicator was developed using data from the Hydrographic Database maintained by the NEFSC Oceans and Climate Branch. Shelf water is defined as all water having salinity <34 PSU (the isohaline typically associated with the shelf-slope front in this region). The methods to compute this time series have been previously published [@fratantoni_description_2013]. 

```{r bsbwsvol, fig.cap="Annual mean winter (February and March) shelf water volume. Dashed line depicts the proposed threshold above which overwinter mixing is more likely.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_winter_shw_vol.R'))
plot_winter_shw_vol()
```


**Golden tilefish Indicators:**

Numerous indicators were developed for the Golden Tilefish ESP [@salois_golden_2024]. Some of those indicators are also relevant to offshore habitat. The indicators pertain to habitat condition and physical oceanography.

Habitat condition indicators assessed the cold pool index, cold pool spatial extent index, cold pool persistence index, and the mean bottom temperature and salinity of the spatial extent in which Golden tilefish have been captured by the NMFS bottom trawl survey. Optimal bottom temperature for golden tilefish is between 9-14 째C [@grimes_tilefish_1986]. Temperatures outside this range may impact spawning location/success and recruitment success. Optimal bottom salinity for juvenile and adult golden tilefish is 33-36 PSU [@steimle_essential_1999]. Salinities outside this range may affect spawning and recruitment success. If the cold pool has a larger extent or persists for a longer time, environmental conditions are not optimal for early life history stages and recruitment success. The cold pool index indicators are generated from bias-corrected ROMS-NWA and GLORYS data. Bottom temperature and salinity indicators were generated using data from GLORYS12 [@eu_copernicus_marine_service_information_global_2023], a global eddy-resolving physical ocean and sea ice model (1993-2023) with high resolution (1/12째).

```{r gtfcp, fig.cap="Cold pool index, cold pool persistence index, and cold pool spatial extent index based on bias-corrected ROMS-NWA (open circles) and GLORYS (closed circles)", echo=FALSE, results = 'hide'}
ecodata::plot_cold_pool()
```


```{r gtfbt, fig.cap="Annual average mean bottom temperature (at 78 m depth) of NMFS bottom trawl stat areas in which Golden tilefish has been captured.Dashed lines depict the upper and lower thermal tolerance of the species. Confidence interval shading represents +/- 2 standard deviations around the weighted mean", echo=FALSE}
source(here::here('offshore_habitat/R/plot_annual_avg_gtf_bottom_temp.R'))
plot_annual_avg_gtf_bottom_temp()
```

```{r gtfbs, fig.cap="Annual average bottom salinity (at 78 m depth) of NMFS bottom trawl stat areas in which Golden tilefish has been captured.Dashed lines depict the upper and lower thermal tolerance of the species. Confidence interval shading represents +/- 2 standard deviations around the average. Trend line represents statistically significant trend", echo=FALSE}
source(here::here('offshore_habitat/R/plot_annual_avg_gtf_bottom_sal.R'))
plot_annual_avg_gtf_bottom_sal()
```


Physical oceanography indicators assessed the sea surface temperature (SST), gulf stream index (GSI), and shelf water volume, temperature, and salinity. SST can be used as a proxy for water mass movement [@fisher_life_2014], which may cause displacement or mortality of eggs and larvae. SST was derived from the Advanced Clear Sky Processor for Ocean (ACSPO) system. A more northerly position of the Gulf Stream (positive GSI) pushes warm water onto the shelf, increasing the amount of suitable habitat for Golden tilefish. The GSI is the degrees latitude above the average Gulf Stream position based on ocean temperature at 200 m (15 C) depth between 55째W to 75째W. Further GSI calculation details are available on the ecodata repository^[https://noaa-edab.github.io/ecodata/reference/gsi.html]. Shelf water volume, temperature, and salinity were derived from the Hydrographic Database maintained by the NEFSC Oceans and Climate Branch [@fratantoni_description_2013]. Shelf water physical oceanography indicators were compare to tolerance thresholds from the literature to assess changes in habitat suitability.

```{r ltsst, fig.cap="Northeast US annual sea surface temperature (SST, black), with increasing trend (orange).", echo=FALSE}
ecodata::plot_long_term_sst()
```
```{r gsi, fig.cap="Gulf stream index. Dashed line depicts time series mean. Trend line depicts statistically significant trend.", echo=FALSE}

ecodata::plot_gsi()

```

```{r gtfshwvol, fig.cap="Annual mean shelf water volume. Dashed line depicts time series mean", echo=FALSE}
source(here::here('offshore_habitat/R/plot_shw_vol.R'))
plot_shw_vol()
```

```{r gtfshwt, fig.cap="Annual mean shelf water temperature. Dashed line depicts time series mean. Trend line depicts statistically significant trend.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_shw_temp.R'))
plot_shw_temp()

```

```{r gtfshws, fig.cap="Annual mean shelf water salinity. Dashed line depicts time series mean.", echo=FALSE}
source(here::here('offshore_habitat/R/plot_shw_sal.R'))
plot_shw_sal()

```





**Risk criteria:**

  Bluefish
  -----------------------------------------------------------------------
  *Risk Level*    *Definition*
  --------------- -------------------------------------------------------
  Low             No indicators suggest negative impact to OY

  Low-Moderate    Either distribution indicators or natural mortality 
                  indicators suggest negative impact to OY but 
                  climate indicators suggest no impact.

  Moderate-High   Climate indicators suggest negative impacts 
                  to OY or Both distribution and natural mortality 
                  indicators suggest negative impacts to OY.

  High            Distribution, climate, and natural mortality 
                  indicators all suggest negative impacts to OY.
  -----------------------------------------------------------------------
  
  Black sea bass  
  -----------------------------------------------------------------------
  *Risk Level*    *Definition*
  --------------- -------------------------------------------------------
  Low             No indicators suggest negative impact to OY

  Low-Moderate    Indicators suggest increase in overwinter mixing
                  but no change or increase in overwinter survival

  Moderate-High   Indicators suggest decrease in overwinter survival
                  but no change or decrease in overwinter mixing

  High            Overwinter survival decreasing, overwinter mixing 
                  increasing
  -----------------------------------------------------------------------
  
  Golden tilefish  
  -----------------------------------------------------------------------
  *Risk Level*    *Definition*
  --------------- -------------------------------------------------------
  Low             No indicators suggest negative impact to OY

  Low-Moderate    Physical oceanography indicators suggest negative
                  impacts to OY but Habitat condition indicators 
                  suggest no impact or positive impact

  Moderate-High   Habitat condition indicators suggest negative 
                  impacts to OY but Physical oceanography indicators
                  suggest no impact or positive impact

  High            Both Physical oceanography and Habitat condition 
                  indicators suggest negative impact to OY
  -----------------------------------------------------------------------

**Risk Assessment**

Bluefish: A mix in indicator trends results in a moderate-high ranking for this element. Climate indicators show mixed signals. The proportion of the central Atlantic warmer than 25.6 C in July, which is negatively correlated with recruitment [@tyrell_bluefish_2022], is increasing. Other climate indicators suggest lengthening spawning season and potentially greater spawning success (increase in last day warmer than 18 C, increase in number of days warmer than 18 C, decrease in proportion of the central Atlantic colder than 18 C in July). Distribution indicators suggest a negative impact to OY as there are significant trends in Medium (30.3-50 cm) and Large (>50 cm) center of gravity both in Eastings and Northings. Natural mortality indicators suggest a positive impact on OY. Mako shark B/Bmsy has a negative trend, indicative of decreased predation and natural mortality of Bluefish. There is a positive trend in Large Bluefish relative condition in the spring which could have positive impacts on survival and reproductive success.

Black sea bass: No indicators suggest a negative impact on OY, resulting in a low ranking for this element. Mean winter bottom temperature is increasing away from the lower temperature threshold of the species which has a positive impact on recruitment. There is no trend in winter mean shelf volume and the majority of the time series is below the volume threshold for increased risk of overwinter mixing between sub regions.

Golden tilefish: No indicators suggest a negative impact on OY, resulting in a low ranking for this element. Trends in the cold pool index and spatial extent suggest warmer conditions and a smaller cold pool which will likely have positive impacts on Golden tilefish as the average cold pool temperature is below the lower thermal tolerance of the species. Increasing trends in SST, GSI, and mean shelf water temperature will similarly result in increases in thermal habitat and therefore survival rates.